An Extension of LTL with Rules\\ and its Application to Runtime Verification

### 2019-03-24 rule-ltl.tex

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the in-expressiveness of transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding alternating quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say B\"{u}chi automata.

The extensions we proposed to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated immediately when this happens, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (publicly available on {\sf github},
hidden here because of anonymous reviewing)
\ignore{(https://github.com/havelund/dejavu)} that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by BDD representation of enumerations of data (rather than over the data values themselves) and using
a garbage collection algorithm.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we also present RV algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

@@

\vspace{1ex}
\noindent {\bf Related work.}
For propositional LTL, the logic ETL~\cite{WVS,Wolper} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator,
where the input letters of the automaton correspond to future temporal subformulas. Each accepted run defines then the places
where the subformulas (corresponding to the occurrences of the input letters in the run) need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

Addition of relations to temporal databases is done
for aggregation~\cite{Libkin}, where relations are
calculated, based on the current state, and are used for further calculating
functions (sums, maxima) over the sequence of states. Aggregations were also used in the runtime verification tool 
{\sf MonPoly}~\cite{agrebasin}.
%
% --- other less directly related work: ---
%
Numerous other systems have been produced for monitoring
execution traces against formal specifications. 
These include e.g. \cite{Meredith2011,larva,Reger2015,halle-beepbeep-ieee-12,LOLA}.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second-order) Q{\em uantification} that is dynamic (state-dependent). This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

@@

Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in PQLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the initial state. Then $q$ alternates between even and odd states.

@@

ELTL extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write conflicting rules, e.g., by using the rule $a_1 : = \false$.

@@

We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
a formalism that is equivalent to finite automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\begin{lemma}
\label{simplecase}
The expressive power of 
EPLTL is the same as QPLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part, may provide a more intuitive specification.
The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}. An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness

@@

The algorithm for PLTL, first presented in \cite{HR}, 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification for a PLTL formula $\eta$ exploits the fact that
the semantic definition is recursive in both the structure of the property. Thus, subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
subformulas for the execution prefix that
ends just {\em before} the new state, and $\current$, for the execution prefix that
ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$
of $\eta$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.

@@

Under {\em mixed evaluation order}, one calculates $\current$ in Step~4 in the following order. The first~4 steps are for the rules, and step~5 is for the statement $\eta$.
\begin{enumerate}
\item Calculate 
$\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j := \varphi_j$, but {\em not} within the scope of
a $\ominus$ operator. (This is not
a problem, since
$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
\item Set 
$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j := \varphi_j$ {\em  within} the scope of
a $\ominus$ operator.
\item Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
\item Calculate $\current ( \eta )$ for the statement
$\eta$.
\end{enumerate}

@@

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
last parameter
that is a natural number represented a time instance. So a tuple of
a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
the $3$ is the time parameter. The last parameter $i$
does not correspond to modeling physical time. This
is rather used to allow the relations to have
different tuples in different instances of $i$,
corresponding to states, in the propositional temporal
logics.

@@

\vspace{1ex}
\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
that in this case, for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015} can express this property, 
where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.

@@

\begin{lemma} \label{samesame}
The auxiliary temporal relations $r_1, \ldots r_n$ are
uniquely defined by the rules for each model ${\cal R}$.
\end{lemma}
\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
\vspace{1ex}
The following formula expresses the property that was shown to be 
not expressible using FLTL, at the beginning of this section.
\begin{equation}
\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
\label{eq:wolper-first-order}
\end{equation}

@@

\noindent 
Analogously to the propositional case, it is easy to show that EPFLTL does not
need the past temporal operators besides $\ominus$.

@@

%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
%and QLTL (PQLTL, respectively) have the same
expressive power. We study now the relationship between
%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).

@@

\noindent {\bf Sketch of Proof.}
The proof of this theorem includes encoding of a  property that
observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
in between states where $r$ appears. The domain of data elements is unbounded.
%(i.e., $r$ is a relation with
%one parameter: the time/state-index parameter $i$).
The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.
The property asserts that there are no two consecutive data sets that are equivalent.
This property can be expressed in QPFLTL. The details appear in the appendix.

>>>>

We then proceed to first-order LTL. We relativise Wolper's example to the first
order case to motivate the need for
extending first-order LTL. Another example is based on the unexpressiveness of the transitive closure of a relation in first order logic, where, e.g., using a relation representing neighbors in a graph is not enough for representing that there is a path between nodes.
We present an extension where auxiliary relations, not appearing in the execution, are
added. While this extension
is capable of expressing the above
properties, we show that for the first-order version, this has less 
expressive power than adding full quantification over relations. This in distinction from adding quantification to propositional LTL, where prefixing existential quantification
provides the full expressive power of, say, B\"{u}chi automata.

The extensions we proposed to the logics are in particular natural for
runtime verification of past (safety) temporal properties, since
the values of propositional and first-order LTL are based
on calculating and updating a summary of the prefix of the execution. Runtime verification is often restricted to a {\em past}  version of temporal logic~\cite{HR}, mainly because for past properties we can decide when the monitored property is violated after monitoring a finite prefix of it, while for future properties we may never know for sure. Consistent with that, 
we provide runtime verification algorithms for the past-time versions of the extended propositional and first-order
logics. We further present an extension of the 
\dejavu{} tool (publicly available on {\sf github},
hidden here because of anonymous reviewing)
\ignore{(https://github.com/havelund/dejavu)} that
realizes the extended first-order past temporal logic verification. The \dejavu{} tool~\cite{HPU,HP} allows runtime verification on past time first-order temporal logic over infinite domains (e.g., the integers, strings). It achieves efficiency by using BDD representation of enumerations of data, rather than over the data values themselves, and by using
a garbage collection algorithm.

The contributions of this paper are both theoretical and practical. On the theory side, we present and study extensions for propositional and first order linear-temporal logics and show the relations between them and existing versions. 
On the practical side, we present RV algorithms for the extended logic, and 
show how to cast them specifically in an efficient BDD-based implementation. We present some experiments performed with our implementation and report on the results.

@@

\vspace{1ex}
\noindent {\bf Related work.}
For propositional LTL, the logic ETL~\cite{Wolper,WVS} extends
the set of temporal operators 
by using automata (or, equivalently a right linear grammar). Each automaton defines a temporal operator, where the input letters of that automaton correspond to future temporal subformulas. Each accepted run of such an automaton defines then the places
where the subformulas need to hold. Another expressive-equivalent extension, 
in~\cite{Wolper}, adds dynamic, i.e.,
state-dependent, quantification over propositions to LTL.

Addition of relations to temporal databases is done
for aggregation~\cite{Libkin}, where relations are
calculated, based on the current state, and are used for further calculating
functions (sums, maxima) over the sequence of states. Aggregations were also used in the runtime verification tool 
{\sf MonPoly}~\cite{agrebasin}.
%
% --- other less directly related work: ---
%
Numerous other systems have been produced for monitoring
execution traces eith data against formal specifications. 
These include e.g. \cite{LOLA,larva,halle-beepbeep-ieee-12,Meredith2011,Reger2015}.

@@

\begin{itemize}
\item{P} restricts to P{\em ast} temporal operators.
\item{F} allows F{\em irst-order} quantification over
data assigned to variables that is uniform for all states. This quantification is referred to as {\em static} quantification.
\item{Q} adds (second-order) Q{\em uantification} that is state-dependent. This is referred to as {\em dynamic} quantification.
\item{E} allows E{\em xtending} the 
set of objects that occur in the input with
new auxiliary predicates (for the propositional versions) or relations (for the first-order cases) using past temporal logic rules that define them.
\end{itemize}

@@

Restricting QPLTL to the past modalities, one obtain the logic PQLTL. PQLTL has the same expressive
power as regular expressions and finite automata.
Equation~(\ref{form0}) can
be rewritten in PQLTL as:
\begin{equation} \label{form1} \Exists q \,
{\bf H} (( q \leftrightarrow \ominus \neg q )   \wedge (q \rightarrow p ) )\end{equation}
%In fact, it can be shown that a single existential quantifier is sufficient. 
Since $\ominus \varphi $ is interpreted as  $\false$ in the first state of any sequence, regardless of $\varphi$, then
$q$ is $\false$ in the first state. Then $q$ alternates between even and odd states.

@@

ELTL extends the set of propositional variables with new variables, whose values at a state are
functions of (i.e., uniquely defined by) the prefix of the model up to that state. The added propositions effectively summarize the past execution prefix up to and
including the current state.
This differs from 
the use of auxiliary variables in QLTL, where the values assigned to the
auxiliary variables do not have to extend the 
states of the model in a unique way throughout the interpretation of 
the property over a model. The constraint that auxiliary variables appearing in the formulas $\varphi_i$ 
must occur within the scope of a $\ominus$ operator
is required to help preventing that the values
of the auxiliary variables are conflicting, as in
$a_1 := \neg a_2$ and $a_2 := a_1$, which makes a formula
that uses $a_1$ and $a_2$ uniformly $\false$. Still, one can write unsatisfiable rules, e.g., by using the rule $a_1 : = \false$.

@@

We define EPLTL by disallowing the future temporal operators (except an implied prefixing $\Box$) in ELTL. This results in
a formalism that is equivalent to finite (B\"{u}chi) automata, where all the states except one are accepting and where the non-accepting state is a sink. 
We can use a related, but simpler construction than in Theorem~\ref{theo1} to prove the following:

@@

\begin{lemma}
\label{simplecase}
The expressive power of 
EPLTL is the same as QPLTL.
%Any safety QPLTL property can be expressed using the following restricted
%ELTL form:
%\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
% {j \in \{1, \ldots , n\}}  \] 
% with $p$ being one of the auxiliary variables %$a_j$.
\end{lemma}
In fact, it is easy to show that EPLTL can be restricted to the following form, while retaining its expressive power.
\[  \Box p {\ where\ } a_j  = \varphi_j \, : \, 
{j \in \{1, \ldots , n\}}  \] 
with $p$ being one of the auxiliary variables $a_j$. Yet, this may not be the preferred form
of an EPLTL property, and the use of further past-time temporal operators, and a non trivial statement  part may provide a more intuitive specification.
The relative expressive power between the propositional temporal logics discussed in this section is summarized in Figure~\ref{prop}. An arrow from A to B means that logic  B is more expressive than A. A double headed arrow shows same expressiveness

@@

The algorithm for PLTL first presented in \cite{HR} 
is based on the observation that the semantics of the 
past time formulas $\ominus \varphi$ and $(\varphi\, \since \,\psi)$ in the current step $i$ is defined in terms of the semantics
in the previous step $i - 1$ of a subformula.
To demonstrate this, we repeat here the definition of the previous-time
operator $\ominus$, and rewrite the definition
of $\since$ in an equivalent form that is
more directly applicable for runtime verification.
\begin{itemize}
\item $( \sigma , i) \models \ominus \varphi$ iff $i > 1$ and $(\sigma, i-1) \models \varphi$.
\item $(\sigma , i) \models (\varphi \, \since \, \psi)$ iff $(\sigma , i) \models \psi$ or the following hold: $i>1$,
$( \sigma , i)  \models \varphi$ and 
$(\sigma , i-1 ) \models ( \varphi \, \since \, \psi )$.
\end{itemize}
Runtime verification for a PLTL formula $\eta$ exploits the fact that
the semantic definition is recursive in both the structure of the property. Thus, subformulas
are evaluated based on smaller subformulas, and the
evaluation of subformulas in the previous state.
The algorithm, shown below, uses two vectors (arrays) of values indexed by subformulas:  $\old$, which summarizes the truth values of the
subformulas for the execution prefix that
ends just {\em before} the new state, and $\current$, for the execution prefix that
ends with the {\em current} state. The order of calculating $\current$ for subformulas is bottom up, according to the syntax tree.
\begin{enumerate}
\item Initially, for each subformula $\varphi$
of $\eta$,
%$B ( \varphi , \current ) = \bfalse $.
$\current ( \varphi ) := \false$.

@@

Under {\em mixed evaluation order}, one calculates $\current$ as part of Step~4 of the above algorithm in the following order. The first~four steps, $a$-$d$, are for the rules, and the fifth step $e$ is for the statement $\eta$.
\begin{description}
\item{\em a.} Calculate 
$\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j := \varphi_j$, but {\em not} within the scope of
a $\ominus$ operator. (This is not
a problem, since
$\current ( \ominus \gamma )$ is set to $\old ( \gamma )$.)
\item{\em b.} Set 
$\current ( a_j )$ to $\current ( \varphi_j )$ for each $j$.
\item{\em c.} Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in $\varphi_j$ of a rule
$a_j := \varphi_j$ {\em  within} the scope of
a $\ominus$ operator.
\item{\em d.} Calculate $\current ( \delta )$ for each subformula $\delta$
that appears in the statement $\psi$, using the calculated $\current ( a_j )$.
\item{\em e.} Calculate $\current ( \eta )$ for the statement
$\eta$.
\end{description}

@@

We define models for FLTL based on {\em temporal} relations~\cite{Chomicki}. That is relations with
last parameter
that is a natural number represented a time instance. So a tuple of
a relation $R$ can be $(``a", 5, ``cbb", 3)$, where
the  value $3$ is the time parameter. The last parameter $i$
does not correspond to modeling physical time. This
is rather used to allow the relations to have
different tuples in different instances of $i$,
corresponding to states, in the propositional temporal
logics.

@@

\vspace{1ex}
\noindent {\bf Example 1.}
Let $p$ and $q$ be temporal relations, where their
restrictions to the state $s_i$, $p[i]$ and $q[i]$, are unary relations $p$ and $q$. The
specification that we want to monitor is that for each value
$a$, $p(a)$ appears in all the states where $q(a)$ has appeared an even number of times so far (for the odd occurrences, $p(a)$ can also appear, but does not have to). To show that this is not expressible in FLTL (and PFLTL),
consider models (executions) where only one data element $a$ appears. We can prove by a structural induction on subformulas
that in this case, for each property $\varphi$
we can replace occurrences of variables within
predicates by $a$, i.e., $p(x)$ and $q(x)$ become $p(a)$
and $q  (a )$, respectively, and that we can throw away
the quantification ($\forall x$, $\exists x$), obtaining
an equivalent formula. This is then equivalent to
an LTL formula that is obtained by replacing
the occurrences of $p(a)$ and $q(a)$ by propositions
$p_a$ and $p_b$, which reduces to Wolper's example~\cite{Wolper}.
Using parametric automata as a specification formalism, as in~\cite{Grum,havelund-rv-data-2018,Meredith2011,Reger2015}, can express this property, 
where for each value $a$ there is a separate automaton that counts the number of times that $q(a)$ has occurred.

@@

@@

\begin{lemma} \label{samesame}
The auxiliary temporal relations $r_1, \ldots r_n$ are
uniquely defined by the rules for each model ${\cal R}$.
\end{lemma}
\noindent {\bf Proof.} By a simple induction, similar to Lemma~\ref{fourone}. \qed
\vspace{1ex}
The following formula expresses the property described in Example~1, which was shown to be 
not expressible using FLTL.
\begin{equation}
\Box \forall x \, (r(x)\rightarrow p(x)) \mathrm{\ where\ }
r(x) = ( q(x) \leftrightarrow \ominus \neg  r(x)) 
\label{eq:wolper-first-order}
\end{equation}
The property that corresponds to Example~2 appears as the third checked example in the implementation section~\ref{implement}.

Analogously to the propositional case, it is easy to show that EPFLTL does not
need the past temporal operators besides $\ominus$.

@@

%We showed in Theorem~\ref{theo1} that ELTL (EPLTL)
%and QLTL (PQLTL, respectively) have the same
%expressive power. We study now the relationship between
%EFLTL (EPFLTL) and QFLTL (QPFLTL, respectively).

@@

\noindent {\bf Sketch of Proof.}
The proof of this theorem includes encoding of a  property that
observes sets of data elements $a$, appearing separately, one per state as $v (a)$,
in between states where $r$ appears. The domain of data elements is unbounded.
%(i.e., $r$ is a relation with
%one parameter: the time parameter $i$).
The set of $a$-values observed in between two consecutive $r$'s is called a {\em data set}.
The property asserts that there are no two consecutive data sets that are equivalent.
This property can be expressed in QPFLTL. The details appear in the appendix.
